### 2015-04-26 papers/histogram/thesis/thesis.tex

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short range forces to first
    order. The infinite potential at $r<\sigma$ simply enforces the
    condition that spheres cannot overlap. At $r>\lambda\sigma$, the
    potential is zero, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
unbiased Monte Carlo) to sufficiently sample $R$, that is, accumulate
a statistically significant histogram $H\p{R}$, in any reasonable
amount of time.

@@

Given the expression for $\bk{X}_T$ in (\ref{eq:XT}), sufficient
accumulation of statistics on $\bk{X}_E$ at all available energies in
principle allows one to determine $\bk{X}_T$ for any temperature
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $\bk{X}_T$ to a reasonable degree of accuracy
requires sufficiently sampling the energies at which $\tilde
D\p{E}e^{-E/kT}$ dominates the sum in (\ref{eq:XT}).

@@

\subsection{The multi-canonical (flat histogram) method}
\label{sec:flat_historam}

The multi-canonical histogram method assumes complete knowledge of the
density of states $D\p E$ of the system in question, and solves
(\ref{eq:dos}) for a weight array $w\p{E}$ which should yield a flat
energy histogram $H\p E=H_0$ to get
\begin{align}
  w\p E=\f1{D\p E},
  \label{eq:flat_weights}
\end{align}
where we may neglect the constant scale factor $H_0$ and do not worry
about normalization factor distinguishing $D\p{E}$ from $\tilde
D\p{E}$ due to the scale-invariance of $w\p{E}$.

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. Nonetheless, this toy method has pedagogical value by
enabling discussion of ``optimal'' weights generate a flat
histogram. All algorithms to construct $w\p{E}$ with the aim of
producing a flat histogram should therefore converge on the same
weight array, given by (\ref{eq:flat_weights}).

\fixme{Talk about a simple implementation of a flat histogram method
  (essentially a single iteration of Simple Flat), and why one might
  want to use it.}

>>>>

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.47\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short range forces to first
    order. The infinite potential at $r<\sigma$ simply enforces the
    condition that spheres cannot overlap. At $r>\lambda\sigma$, the
    potential is null, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$; that is, the number of observations
$H\p{X_0}$ of the system with $X\p{s}=X_0$ is proportional to the
total number of states $s$ for which $X\p{s}=X_0$. It is sometimes the
case, however, that the density of states in some region $R$ in the
range of possible $X$ is so low that it is practically impossible (via
unbiased Monte Carlo) to sufficiently sample $R$, that is, accumulate
a statistically significant histogram $H\p{X\in R}$, in any reasonable
amount of time.

@@

Given the expression for $\bk{X}_T$ in (\ref{eq:XT}), sufficient
accumulation of statistics on $\bk{X}_E$ at all available energies in
principle allows one to determine $\bk{X}_T$ for any temperature
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $\bk{X}_T$ to a reasonable degree of accuracy
requires sufficiently sampling the energies at which $\tilde
D\p{E}e^{-E/kT}$ dominates the sum in (\ref{eq:XT}). Section
\ref{sec:energy_range} provides more discussion on this subject,
namely on identifying the energies which are important for computing
system properties at a given temperature.

@@

\subsection{The flat histogram (multi-canonical) method}
\label{sec:flat_historam}

The flat histogram method, also called the multi-canonical method,
assumes complete knowledge of the density of states $D\p{E}$ of the
system in question, and solves (\ref{eq:dos}) for a weight array
$w\p{E}$ which should yield a flat energy histogram $H\p E=H_0$ to get
\begin{align}
  w\p E=\f1{D\p E},
  \label{eq:flat_weights}
\end{align}
where we may neglect the constant scale factor $H_0$ and do not worry
about normalization factor distinguishing $D\p{E}$ from $\tilde
D\p{E}$ due to the scale-invariance of $w\p{E}$.

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. A naive way to implement the flat histogram method without
knowing a priori the density of states $D\p{E}$ involves first running
a simulation without weights (or with constant weights $w\p{E}=w_0$)
while collecting a histogram $H\p{E}$ of the observed energies
$E$. After some time (loosely defined), the histogram $H\p{E}$ may be
taken as an approximation of the density of states $\tilde D\p{E}$
(see Section \ref{sec:histogram_methods}), and used via
(\ref{eq:flat_weights}) to generate a weight array for an actual
simulation.

The problem with this implementation is that histogram methods are
generally necessary when some range of energies $R$ is inaccessible
via unbiased Monte Carlo simulations. In such a case, the energy
histogram $H\p{E\in R}$ after simulating for any reasonable amount of
time will be statistically insignificant or null, invalidating the
approximation $H\p{E}\approx\tilde D\p{E}$. This method will therefore
only ``flatten'' the histogram at energies $E$ which have been
sufficiently sampled for $H\p{E}$ to be statistically significant. The
end goal in implementing histogram methods, however, is precisely to
sample those energies that cannot be sufficiently sampled without
employing clever schemes to bias Monte Carlo simulations. The flat
histogram method therefore does not directly help us study the
square-well fluid. Nonetheless, the this method has pedagogical value
by enabling discussion of an ``optimal'' weight array for generating a
flat energy histogram in simulations. All algorithms to construct
$w\p{E}$ with the aim of producing a flat energy histogram should
converge on the same weight array, given by (\ref{eq:flat_weights}).
